---
title: "Math Primer: Difference Equations and Ordinary Differential Equations (ODEs)"
output: webexercises::webexercises_default
---

```{r setup, include=FALSE}
library(webexercises)
knitr::opts_chunk$set(echo = FALSE)
```

## The mathematics of change - difference equations and ordinary differential equations (ODEs).

This tutorial provides a brief introduction to the these two fundamental types of equations that are used to model change in many fields of science and engineering:

-   *Ordinary differential equations* are used to model situations where the change is assumed to occur continuously over time or space. In the context of time, these equations relate the instantaneous rate of change of a system to its current state. This is formally expressed in the form of derivatives (e.g. $\tfrac{dy}{dt}=t^{2}$).
-   *Difference equations* are used to model situations where the change can be assumed to occur at distinct points in time or space. In the context of time, these equations link the state of a mathematical system at the present time (often referred to as time $t$ ) to a future time (such as time $t+1$).

These two representations of change are complementary, but the underlying mathematical framework and the tools that we use to manipulate them are different.

### Example difference equations and ODEs for a simple system.

Imagine that we have a biological process that changes over time, such as the depletion of a nutrient by a growing population of bacteria. We can represent this process mathematically as:

$$\frac{dN}{dt}=-kN$$

where $N$ is the (log) concentration of nutrient and $k$ is the rate constant for the depletion of the nutrient. This is an example of an ODE, because the rate of change of the nutrient concentration is assumed to be continuous over time.

Now imagine that we have a similar biological process, but the nutrient is being consumed by a population of bacteria that grows in discrete steps. We can represent this process mathematically as:

$$N_{t+1}=N_{t}-kN_{t}$$

where $N_{t}$ is the concentration of nutrient at time $t$ and $N_{t+1}$ is the concentration of nutrient at the next time step. This is an example of a difference equation, because the rate of change of the nutrient concentration is assumed to be discrete.

Although the dynamics of these systems are not identical (because the rate of change is assumed to be continuous in the ODE and discrete in the difference equation), they are related. In fact, the difference equation can be derived from the ODE by assuming that the rate of change is constant over a small time interval (e.g. $\Delta t$) and then rearranging the terms to give: $$\frac{N_{t+1}-N_{t}}{\Delta t}=-kN_{t}$$

The key idea is that when the time interval $\Delta t$ is relatively small, the difference equation is a good approximation of the ODE. This is illustrated in the figure below, which shows the solution of the ODE (blue line) and the solution of the difference equation (red circles) for the same system with the same initial conditions and parameter values.

```{r difference-equation-ode, fig.width=6, fig.height=4}
# ODE Solver package
library(deSolve)

# Define the ODE (function for the derivative dN/dt).
decay_ode <- function(t, N, k) {
  dN <- -k * N
  return(list(dN))
}

# Define the difference equation
decay_diff <- function(N, k) {
  N_next <- N - k * N
  return(N_next)
}

# Initial conditions
k <- 0.15
t_start <- 0
t_end <- 20
t <- seq(t_start,t_end,0.01)
N0 <- 100

# Numerical solution of the ODE
ode_sol <- ode(y = N0, times = t, func = decay_ode, parms = k)

# Difference equation.
deltaT <- 1
steps <- (t_end - t_start) / deltaT + 1
t_diff <- seq(t_start, t_end, deltaT)
N_out <- rep(NA, steps)
N_out[1] <- N0

for(i in 2:steps){
  N0 <- ifelse(i == 1, N0, N_out[i-1])
  N_out[i] <- decay_diff(N0, k)
}

plot(ode_sol, type = "l", lwd = 2, col = "blue", xlab = "Time", ylab = "N", main="ODE (blue) vs. Difference Equation (red)")
points(t_diff,N_out, pch = 16, col = "red")
```

::: {.webex-check .webex-box}
### Check your understanding

```{r quiz1, results='asis'}
opts <- c(answer = "Difference equations represent discrete change, while ODEs represent continuous change",
                   "Difference equations represent continuous change, while ODEs represent discrete change",
                   "Difference equations are linear, while ODEs are nonlinear",
                   "Difference equations are solvable, while ODEs are not")

cat("What is a key distinction between difference equations and ODEs?", longmcq(opts))

opts2 <- c(answer = "When the time interval is relatively small",
                    "When the time interval is relatively large",
                    "When the time interval is zero",
                    "When the time interval is infinite")
cat("Under what conditions can difference equations be good approximations of an ODE?",longmcq(opts2))
```
:::

## Ordinary Differential Equations and Population Growth - the Exponential and Logistic models

Biological populations grow and decline over time, and the dynamics of these changes can be modeled using ordinary differential equations. In this section, we will explore two of the most common models of population growth - the exponential model and the logistic model. Both of these models are fundamentally nonlinear, reflecting the fact that the rate of population growth is not constant over time. However, they differ in the way that they represent the dynamics of population growth.

### The Exponential Model

The exponential model is the simplest model of population growth, and it is based on the assumption that the rate of growth is proportional to the current population size. The intuition behind this assumption is that adding additional individuals to the population also increases the populations ability to make additional offspring. This assumption is reasonable when the population is small relative to the ability of the environment to support additional individuals, and when the population is growing rapidly. The exponential model is represented mathematically as an ODE:

$$\frac{dN}{dt}=rN$$

where $N$ is the population size and $r$ is the intrinsic rate of population growth. The figure below shows the solution to the exponential model for a population with an initial size of 100 individuals and an intrinsic rate of population growth of 0.15 individuals per time step.

```{r exponential-model, fig.width=6, fig.height=4}
# ODE Solver package
library(deSolve)

# Define the ODE (function for the derivative dN/dt).
exp_ode <- function(t, N, r) {
  dN <- r * N
  return(list(dN))
}

# Initial conditions
r <- 0.15
t_start <- 0
t_end <- 25
t <- seq(t_start,t_end,0.01)
N0 <- 100

# Numerical solution of the ODE
ode_sol <- ode(y = N0, times = t, func = exp_ode, parms = r)

plot(ode_sol, type = "l", lwd = 2, col = "slateblue", xlab = "Time", ylab = "N", main="Exponential Model",
     ylim=c(100,2000))

```

The exponential growth ODE is simple enough to have an [analytical solution](). This means that we can directly solve for the population size $N$ at any time $t$ without having to use a numerical solver. The analytical solution to the exponential growth ODE is:

$$N(t)=N_{0}e^{rt}$$

where $N_{0}$ is the initial population size.

### The Logistic Model

The logistic model is a more realistic model of population growth than the exponential model, because it incorporates the idea that the rate of population growth is limited by the carrying capacity of the environment. The carrying capacity is the maximum population size that can be supported by the environment, and it is determined by the amount of resources available to the population. The logistic model is represented mathematically as an ODE:

$$\frac{dN}{dt}=rN\left(1-\frac{N}{K}\right)$$

where $N$ is the population size, $r$ is the intrinsic rate of population growth, and $K$ is the carrying capacity. The figure below shows the solution to the logistic model for a population with an initial size of 100 individuals, an intrinsic rate of population growth of 0.15 individuals per time step, and a carrying capacity of 1000 individuals.

```{r logistic-model, fig.width=6, fig.height=4}
# ODE Solver package
library(deSolve)

# Define the ODE (function for the derivative dN/dt).
logistic_ode <- function(t, N, parms) {
  r <- parms[1]
  K <- parms[2]
  dN <- r * N * (1 - N / K)
  return(list(dN))
}

# Parameters and Initial conditions
r <- 0.15
K <- 1000
t_start <- 0
t_end <- 25
t <- seq(t_start,t_end,0.01)
N0 <- 100

parms <- c(r=r, K=K)

# Numerical solution of the ODE
logistic_sol <- ode(y = N0, times = t, func = logistic_ode, parms = parms)

# Plots both models
plot(ode_sol, type = "l", lwd = 2, col = "slateblue", xlab = "Time", ylab = "N", main="Exponential vs. Logistic Model",
     ylim=c(100,2000))
points(logistic_sol, type = "l", lwd = 2, col = "orange", xlab = "Time", ylab = "N",
     ylim=c(100,2000))
points(t,rep(K,length(t)), type="l", lty=2, col= "black")
legend("topleft", legend=c("Exponential Model","Logistic Model","Carrying Capacity, K"), col=c("slateblue","orange","black"), lty=c(1,1,2), cex=0.8)

```

The behavior of the logistic model is intuitive: when the population is small relative to the carrying capacity, $K$. The population grows similarly to the exponential model. As the population approaches the carrying capacity, however, population growth slows, and will approach but never reach the value of $K$.

### Adding an Outside Influence (Migration) to the Logistic Model

The models we've reviewed so-far assume a "closed" population, where individuals do not leave or arrive, but in nature this is rarely the case. We can extend the logistic model to include migration by adding another term to the ODE that represents the number of individuals that arrive to the population from outside the population. The ODE for the logistic model with migration is:

$$\frac{dN}{dt}=rN\left(1-\frac{N}{K}\right)+m$$

where $N$ is the population size, $r$ is the intrinsic rate of population growth, $K$ is the carrying capacity, and $m$ is the number of individuals that arrive to the population from outside it. You will note that this is identical to the logistic model above, except for the addition of the migration term. The figure below shows the solution to the logistic + migration model using the same parameters and initial conditions as the original logistic model, with a migration rate of 15 individuals per time step.

```{r logistic-model-migration, fig.width=6, fig.height=4}
# ODE Solver package
library(deSolve)

# Define the ODE (function for the derivative dN/dt).
logistic_migration_ode <- function(t, N, parms) {
  r <- parms[1]
  K <- parms[2]
  m <- parms[3]
  dN <- r * N * (1 - N / K) + m
  return(list(dN))
}

# Parameters and Initial conditions
r <- 0.15
K <- 1000
m <- 15
t_start <- 0
t_end <- 25
t <- seq(t_start,t_end,0.01)
N0 <- 100

parms <- c(r=r, K=K, m=m)

# Numerical solution of the ODE
migration_sol <- ode(y = N0, times = t, func = logistic_migration_ode, parms = parms)

# Plots both models
plot(ode_sol, type = "l", lwd = 2, col = "slateblue", xlab = "Time", ylab = "N", 
     main="Logistic Model with In-migration", ylim=c(100,2000))
points(logistic_sol, type = "l", lwd = 2, col = "orange", xlab = "Time", ylab = "N",
     ylim=c(100,2000))
points(migration_sol, type = "l", lwd = 2, col = "green", xlab = "Time", ylab = "N",
     ylim=c(100,2000))
points(t,rep(K,length(t)), type="l", lty=2, col= "black")
legend("topleft", legend=c("Exponential Model","Logistic Model","Logistic + in-migration", "Carrying Capacity, K"), col=c("slateblue","orange","green","black"), lty=c(1,1,1,2), cex=0.8)
```

Note that the population size in the logistic + migration model is always larger than the logistic model, and approaches the carrying capacity more quickly. This is because the migration term adds individuals to the population, increasing the population size relative to the logistic model. As the population approaches the carrying capacity, the additional migrating individuals will keep the maximum population size above the carrying capacity.

::: {.webex-check .webex-box}
### Check your understanding

```{r quiz2, results='asis'}
opts <- c(answer = "As the size of the population grows, the number of reproductive individuals added to the population each time step increases.",
                  "As the size of the population grows, the number of reproductive individuals added to the population each time step decreases.",
                  "More individuals are attracted to the larger population, increasing in-migration as the population size grows",
                  "The rate of population growth is constant over time in the exponential model")

cat("Why does the rate of population growth accelerate over time in the exponential model?", longmcq(opts))

opts <- c(answer = "The net population growth rate. The number of new individuals born in the population each time step, minus the number of individuals that die.",
                  "The number of new individuals born in the population at each time step.",
                  "The number of new individuals born in the population at each time step, plus the number of individuals that die.",
                  "The number of migrants that arrive in the population each time step.")

cat("In both exponential and logistic models, what does the population growth rate, r, represent?", longmcq(opts))

```
:::

## Difference Equation Approximations for Population Growth

In the above examples, we used ordinary differential equations to model exponential and logistic population growth. However, we can also use difference equation approximations to solve these models. Remember that the difference equation approach is a discrete-time approximation of the continuous-time ODE. When would it be desirable to use a difference equation approximation instead of the ODE? One example is when we have data on the population size at discrete time steps, but not continuous data on the population size. Another example is when we want to simulate the population growth model in a computer program. In this case, we can use the difference equation approximation to simulate the population growth model.

Recall that the derivative of a function over time is defined as:

$$\frac{df}{dt}=\lim_{\Delta t\to 0}\frac{f(t+\Delta t)-f(t)}{\Delta t}$$
This means that as the time interval $\Delta t$ approaches zero, the derivative of the function $f(t)$ is equal to the difference between the function evaluated at time $t+\Delta t$ and the function evaluated at time $t$, divided by the time interval $\Delta t$. We can rearrange this equation to solve for $f(t+\Delta t)$:

$$f(t+\Delta t)=f(t)+\frac{df}{dt}\Delta t$$
This equation is the basis for the difference equation approximation of the ODE. We can approximate the derivative of the function $f(t)$ as the difference between the function evaluated at time $t+\Delta t$ and the function evaluated at time $t$, divided by the time interval $\Delta t$. We can then rearrange the equation to solve for $f(t+\Delta t)$, which is the value of the function at the next time step.

...additional details on difference equation approximations...

### The Logistic Difference Model

We can use the difference equation approximation to solve the logistic model. Recall that the logistic model is defined by the ODE:

$$\frac{dN}{dt}=rN\left(1-\frac{N}{K}\right)$$
...additional details on difference equation approximation...


### Check your understanding

## Exercises for Difference Equations and ODEs
